No preposition with inanimate but an \tsc{r}-pronoun with postposition\\ On \tit{waarmee} vs. \tit{met wat} in Dutch

### 2020-01-14 paper.tex
mid tuesday

\title{From *[P pronoun\scsub{inanim}] to [\tsc{r}-pronoun-P]\\ On \tit{waarmee} in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

@@

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a good insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a proper constituent (i.e. a constituent to the exclusion of other features), the \tsc{r}-pronoun surfaces. If it is not a proper constituent, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. --here more about the internal structure of r-pronouns, and the suffix prefix distinction--

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities and differences between the two forms. First, I identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{'rmee} `with it' and \tit{met 't} `with it'. I decompose these two expressions further, and end up arguing for a picture as shown in Table \ref{tbl:nofeatures}.

The elements \tit{w} and \tit{a} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{mee} and \tit{'r} together, but the distribution differs. \tit{Met} expresses less structure than \tit{mee}, and \tit{'t} expresses more structure than \tit{'r}. \tit{Met} is shown to be an prefix, and \tit{mee} is a suffix. It also becomes clear that the \tit{'r} in fact corresponds to the locative \tit{r} in Dutch.

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \phantom{\tsc{wh}}  & \phantom{\tsc{deix}}                    & \phantom{\tsc{f}4}  & \phantom{\tsc{f}3} & \phantom{\tsc{f}2}  & \phantom{\tsc{f}1}  & \phantom{\tsc{thing}} \\
		%\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{'t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{'r}  \\
    \bottomrule
\end{tabularx}
\label{tbl:nofeatures}
\end{minipage}
\end{table}

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. Exactly because they appear in both expression, I do not pay much attention to them. For \tit{w} I follow \citet{hachem2015} in that \tit{w} realizes a WP.

@@

First of all, when two lexical entries are both candidates for spellout, the most specific is inserted, following the Elsewhere Condition:

@@

In this section I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a suffix, and its form changes into \tit{mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

This section first discusses some basic facts about case and assumptions about Nanosyntax. Then I discuss the distinction between prefixes and suffixes, followed by a discussion on suppletion in pronouns. Finally, I specify the lexical entries for \tit{'rmee} `with it` and \tit{met 't} `with it` that account for the described facts.

\subsubsection{Case and nanosyntax}

Let me continue with specifying the lexical entries for \tit{met 't} `with it'. \tit{'t} `it' can be used as subject, direct object and indirect object, as shown in \ref{ex:tsubobj}.

\ex.\label{ex:tsubobj}
\ag. 't Staat in de hal.\\
 \tsc{3sg.n.nom} stands in the hallway\\
 `It is standing in the hallway.'
\bg. Ik zie 't.\\
 I see \tsc{3sg.n.acc}\\
 `I see it.'
\bg. Ik heb 't een klap gegeven.\\
 I have \tsc{3sg.n.dat} a hit given\\
 `I hit it.'

Pronouns in other genders alternate between nominative (non-oblique) and accusative/dative (oblique) in these contexts, illustrated in \ref{ex:hemsubobj}.

\ex.\label{ex:hemsubobj}
\ag. Hij staat in de hal.\\
 \tsc{3sg.m.nom} stands in the hallway\\
 `He is standing in the hallway.'
\bg. Ik zie hem.\\
 I see \tsc{3sg.m.acc}\\
 `I see it.'
\bg. Ik heb hem een klap gegeven.\\
 I have \tsc{3sg.m.dat} a hit given\\
 `I gave him a smack.'

I assume that the inanimate personal pronoun minimally contains the ontological category \tsc{thing} \citep{kayne2005}. I leave possible number and gender features out because they do not play a role in this paper.

The morpheme \tit{'t} can act as nominative, accusative and dative, as I showed in \ref{ex:tsubobj}. Taking this all together, \tit{'t} should have the lexical entry given in \ref{ex:entryt}.

\ex. \begin{forest} boom
 [\tsc{dat}P
     [\tsc{f}3]
     [\tsc{acc}P
         [\tsc{f}2]
         [\tsc{nom}P
             [\tsc{f}1]
             [\tsc{thingP}
                 [\tsc{thing}, roof]
             ]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{'t}}; }
 \end{forest}\label{ex:entryt}

 IT CANNOT BE STRONG, but it is weak 't staat in de hal

This lexical entry can lexicalize the \tsc{datP}, but also the \tsc{accP} and \tsc{nomP}, due to the Superset Principle.

START WITH 'T IN DUTCH, JUST HOW I DID BEFORE. SUPERSET PRINCIPLE, CASE HIERARCHY, ALL OF IT

THEN INTRODUCE PREFIX/SUFFIX DISTINCTION.

I discuss the basics of spellout and case..

@@

\ex. The preposition/suffix hierarchy
\a. If the expression of a particular case in the Case sequence (below) involves a preposition, then all cases to its right do as well.
\b. The Case sequence: \tsc{nom – acc – gen – dat – ins – com} \hfill \citep{2009}

put this in syntax

now derive it with movements

@@

@@

Nanosyntax distinguishes prefixes and suffixes by distinguishing these types of lexical entries. This is the crucial difference needed for the Dutch instrumental.

It differs per language and within a language how high a DP can move. Everything below is a suffix on the noun, everything higher is a preposition. Consider German full DP. Moves as high as above the dative, instrumental is expressed with a preposition.

@@

\ex.
\ag. Tazi duma m-i e nepoznata.\\
that word I-\tsc{dat} is unfamiliar\\
`That word is unfamiliar to me.'
\bg. Tazi duma e nepoznata sina mi.\\
that word is unfamiliar son my\\
`That word is unfamiliar my son.'
\bg. Tazi duma e nepoznata na sina mi.\\
that word is unfamiliar to son my\\
`That word is unfamiliar to my son.'

The same intuition will hold for Dutch.
-mee is going to be a suffix that will be able to attach to 'r, but it will not to full DPs and animates.

@@

\tit{'r} `there' can be used as a locative.

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.

\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

Notice already here that, via the superset principle, \tit{'r} can be used to realize the feature \tsc{thing} as well, as it is contained in \tsc{locP}.

@@

Note that there are two differences between the lexical entry in \ref{ex:entrymet} and \ref{ex:entrymee}. First, the lexical entry for \tit{mee} `with' in \ref{ex:entrymee}, in addition to the one for \tit{met} `with' in \ref{ex:entrymet}, contains the features \tsc{f1}, \tsc{f2} and \tsc{f3}.
Second, the lexical entry in \ref{ex:entrymet} has a binary bottom, and the lexical entry in \ref{ex:entrymee} has a unary bottom. \tit{Met} `with' is a prefix (preposition), so it has a binary bottom, and \tit{mee} `with' is a suffix, so it needs a unary bottom. In the next section I show how this works in a derivation.

\footnote{For completeness, \tit{mee} `with' has to be blocked from combining with full DPs and animate pronouns (reference). For now I assume that these pronouns spell out an additional feature \tit{σ} that is not realized by \tit{mee} `with' but only by \tit{met} `with'. Full DPs combine in addition with a zero marker for dative case. This way \tit{mee} `with' does not compete with \tit{met} `anymore'. I left this features out in this paper, at is does not contribute to the main point.}

\footnote{Giving \tit{met} `with' and \tit{mee} `with' two distinct lexical entries has as a consequence that the phonological overlap between them seems like a coincidence. Moreover, in almost all cases the preposition does not change form when it combines with an \tsc{r}-pronoun, e.g. \tit{in}. If this proposal is on the right track, elements as \tit{in} can be used as either a prefix and as a suffix. A lexical entry as in \ref{ex:presuf} would be a candidate for such an element.

@@

I leave it to future research to determine whether this is a feasible solution.}

\footnote{A topic related to this paper is the different positioning of identical adpositions in Dutch (see \citet{caha2010} for an account of German and Dutch and \citet{pretorius2017} for Afrikaans). In \ref{ex:dutchin}, \tit{in} changes meaning dependening on whether it proceeds or follows the DP, it is respectively locational or directional.

@@

Skipping over the irrelevant details of how \tit{w} and \tit{a} end up in their positions, the structure for \tit{waarmee} `with what' look as in \ref{ex:spelloutwaarmee}.
\footnote{I assume that WP and \tsc{deixP} are both complex specifiers that are created after \tsc{thing} is spelled out in \ref{ex:thingspellout}. In after each merge after that, backtracking always takes place, the complex specifier is detached from the structure and the case features are spelled out together with or as a suffix on \tsc{thing}.}

@@

A consequence of analyzing \tit{mee} `with' as a suffix is that \tit{r} and \tit{mee} always form a constituent to the exclusion of \tit{w} and \tit{a}. At first sight this seems problematic, because it is possible for \tit{waar} `where', stranding \tit{mee} `with'. I repeat the relevant example in \ref{ex:meestranded}.

@@

The described pattern follows from a core assumptions in Nanosyntax: phrasal spellout spells out constituents. Looking more into detail, \tit{waarmee} `with what' takes precedence over \tit{met wat} `with what' because \tit{mee} has the structure of a suffix. Following the spellout algorithm, \tit{'t} `it' is replaced by \tit{'r} `there', so that it can combine with \tit{mee} `with'. The preposition \tit{met} `with' is only used if there is no other option to spell out the features.

>>>>

\title{No preposition with inanimate but an \tsc{r}-pronoun with postposition\\
On \tit{waarmee} vs. \tit{met wat} in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

In answering the question why [P inanimate] I look at an exception to the rule that *P inanimate pronoun. That exception is a mismatching free relative. Here the string \tit{met wat} does appear.

@@

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a good insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a proper constituent (i.e. a constituent to the exclusion of other features), the \tsc{r}-pronoun surfaces. If it is not a proper constituent, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. --here more about the internal structure of r-pronouns, and the postposition preposition distinction--

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities and differences between the two forms. First, I identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{'rmee} `with it' and \tit{met 't} `with it'.

The elements \tit{w} and \tit{a} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{mee} and \tit{'r} together, but the distribution differs. \tit{Met} expresses less structure than \tit{mee}, and \tit{'t} expresses more structure than \tit{'r}. \tit{Met} is shown to be an preposition, and \tit{mee} is a postposition. It also becomes clear that the \tit{'r} in fact corresponds to the locative \tit{r} in Dutch.

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. Exactly because they appear in both expression, I do not pay much attention to them. For \tit{w} I follow \citet{hachem2015} in that \tit{w} realizes a WP.\footnote{Throughout the paper, ⇔ indicates the pairing between a lexical tree and a phonological form in a lexical entry, and ⇒ indicates how a node in the syntactic structure is spelled out.}

@@

@@

In this section I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition, and its form changes into \tit{mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

This section first discusses some basic facts about case and assumptions about Nanosyntax. Then I discuss the distinction between prepositiones and postpositiones, followed by a discussion on suppletion in pronouns. Finally, I specify the lexical entries for \tit{'rmee} `with it` and \tit{met 't} `with it` that account for the described facts.

\subsubsection{\tit{'t} vs. \tit{'r}}

Let me start with the lexical entry for \tit{'t}. \tit{'t} `it' can be used as subject (associated with nominative), direct object (associated in accusative) and indirect object (associated with dative), as shown in \ref{ex:tsubobj}.

\ex.\label{ex:tsubobj}
\ag. 't Staat in de hal.\\
 \tsc{3sg.n.nom} stands in the hallway\\
 `It is standing in the hallway.'\label{ex:tnoclitic}
\bg. Ik zie 't.\\
 I see \tsc{3sg.n.acc}\\
 `I see it.'
\bg. Ik heb 't een klap gegeven.\\
 I have \tsc{3sg.n.dat} a hit given\\
 `I gave it a hit.'

 Following the distinctions from \citet{cardinaletti1996}, \tit{'t} `it' is a weak pronoun. It is not a clitic, because it can occur in sentence initial position, shown in \ref{ex:tnoclitic}. It is not a strong pronoun, because it cannot be coordinated, as indicated in  \ref{ex:tcoordinated}. \ref{ex:datcoordinated} shows that \tit{'t} `it' needs to combine with \tit{da-}/\tit{di-} to be able to be coordinated.

 \ex.
 \ag. *Hij en 't staan in de hoek.\\
  he and it stand in the corner\\
  `He and it are standing in the corner.'\label{ex:tcoordinated}
 \bg. Hij en dit/dat staan in de hoek.\\
  he and this/that stand in the corner\\
  `He and it are standing in the corner.'\label{ex:datcoordinated}

I assume that the \tit{'t} contains the ontological category \tsc{thing} \citep{kayne2005}. The feature Σ indicates that the pronoun is a weak pronoun. I leave possible number and gender features out because they do not play a role in this paper. The morpheme \tit{'t} can act as nominative, accusative and dative, as I showed in \ref{ex:tsubobj}. Taking this all together, \tit{'t} has the lexical entry given in \ref{ex:entryt}.

\ex. \begin{forest} boom
 [\tsc{dat}P
     [\tsc{f}3]
     [\tsc{acc}P
         [\tsc{f}2]
         [\tsc{nom}P
             [\tsc{f}1]
             [ΣP
                 [Σ]
                 [\tsc{thingP}
                     [\tsc{thing}, roof]
                 ]
             ]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{'t}}; }
 \end{forest}\label{ex:entryt}

 This lexical entry can lexicalize the \tsc{datP}, but also the \tsc{accP} and \tsc{nomP}. This is due to the Superset Princple.

 In other words, a lexically stored structure does not have to be identical to the syntactic structure. It is enough if the syntactic structure is contained in the lexically stored tree. This has a as consequence that the lexical entry in \ref{ex:entryt} can also be inserted in \ref{ex:tacc} and \ref{ex:tnom}.

 \ex.
 \a. \begin{forest} boom
 [\tsc{acc}P
     [\tsc{f}2]
     [\tsc{nom}P
         [\tsc{f}1]
         [\tsc{thingP}
             [\tsc{thing}, roof]
         ]
     ]
 ]
 \end{forest}\label{ex:tacc}
 \b. \begin{forest} boom
 [\tsc{nom}P
     [\tsc{f}1]
     [\tsc{thingP}
         [\tsc{thing}, roof]
     ]
 ]
 \end{forest}\label{ex:tnom}

Let me move on to \tit{'r}. \tit{'r} `there' can be used as a locative.

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

Notice already here that, via the superset principle, \tit{'r} can be used to realize the feature \tsc{thing} as well, as it is contained in \tsc{locP}. Moreover, in a syntactic structure like in \ref{ex:rthing} the lexical entry \ref{ex:entryr} will be inserted and not \ref{ex:entryt}.

\ex.
\begin{forest} boom
 [\tsc{thingP}
     [\tsc{thing}, roof]
 ]
{\draw (.east) node[right]{⇒ \tit{'r}}; }
\end{forest}\label{ex:rthing}

This is due to the Elsewhere condition. The idea is that when two lexical entries are both candidates for spellout, the most specific is inserted.

The syntactic structure in \ref{ex:entryr} only has \tsc{loc} as an unused feature, whereas \ref{ex:entryt} has Σ up to \tsc{f3}.

\subsubsection{\tit{Mee} vs. \tit{met}}

\ex. The preposition/postposition hierarchy
\a. If the expression of a particular case in the Case sequence (below) involves a preposition, then all cases to its right do as well.
\b. The Case sequence: \tsc{nom – acc – gen – dat – ins – com} \hfill \citep{2009}

@@

There is variation, between languages but also within languages.

\ex.
\ag. Tazi duma m-i e nepoznata.\\
that word I-\tsc{dat} is unfamiliar\\
`That word is unfamiliar to me.'
\bg. *Tazi duma e nepoznata sina mi.\\
that word is unfamiliar son my\\
`That word is unfamiliar my son.'
\bg. Tazi duma e nepoznata na sina mi.\\
that word is unfamiliar to son my\\
`That word is unfamiliar to my son.'

show lexical entry for this?

Here pronouns can take a postposition, but full DPs cannot. In Dutch it was slightly different: inanimate pronouns can take the postposition, but animates and full DPs cannot.

@@

The distinction between prepositions and postpositions.

@@

Nanosyntax distinguishes prepositiones and postpositiones by distinguishing these types of lexical entries. This is the crucial difference needed for the Dutch instrumental.

It differs per language and within a language how high a DP can move. Everything below is a postposition on the noun, everything higher is a preposition. Consider German full DP. Moves as high as above the dative, instrumental is expressed with a preposition.

@@

The same intuition will hold for Dutch.
-mee is going to be a postposition that will be able to attach to 'r, but it will not to full DPs and animates.

@@

@@

Note that there are two differences between the lexical entry in \ref{ex:entrymet} and \ref{ex:entrymee}. First, the lexical entry for \tit{mee} `with' in \ref{ex:entrymee}, in addition to the one for \tit{met} `with' in \ref{ex:entrymet}, contains the features \tsc{f1}, \tsc{f2} and \tsc{f3}.
Second, the lexical entry in \ref{ex:entrymet} has a binary bottom, and the lexical entry in \ref{ex:entrymee} has a unary bottom. \tit{Met} `with' is a preposition (preposition), so it has a binary bottom, and \tit{mee} `with' is a postposition, so it needs a unary bottom. In the next section I show how this works in a derivation.\footnote{For completeness, \tit{mee} `with' has to be blocked from combining with full DPs and animate pronouns (reference). For now I assume that these pronouns spell out an additional feature \tit{σ} that is not realized by \tit{mee} `with' but only by \tit{met} `with'. Full DPs combine in addition with a zero marker for dative case. This way \tit{mee} `with' does not compete with \tit{met} `anymore'. I left this features out in this paper, at is does not contribute to the main point.}$^,$\footnote{Giving \tit{met} `with' and \tit{mee} `with' two distinct lexical entries has as a consequence that the phonological overlap between them seems like a coincidence. Moreover, in almost all cases the preposition does not change form when it combines with an \tsc{r}-pronoun, e.g. \tit{in}. If this proposal is on the right track, elements as \tit{in} can be used as either a preposition and as a postposition. A lexical entry as in \ref{ex:presuf} would be a candidate for such an element.

@@

I leave it to future research to determine whether this is a feasible solution.}$^,$\footnote{A topic related to this paper is the different positioning of identical adpositions in Dutch (see \citet{caha2010} for an account of German and Dutch and \citet{pretorius2017} for Afrikaans). In \ref{ex:dutchin}, \tit{in} changes meaning dependening on whether it proceeds or follows the DP, it is respectively locational or directional.

@@

Skipping over the irrelevant details of how \tit{w} and \tit{a} end up in their positions, the structure for \tit{waarmee} `with what' look as in \ref{ex:spelloutwaarmee}.\footnote{I assume that WP and \tsc{deixP} are both complex specifiers that are created after \tsc{thing} is spelled out in \ref{ex:thingspellout}. In after each merge after that, backtracking always takes place, the complex specifier is detached from the structure and the case features are spelled out together with or as a postposition on \tsc{thing}.}

@@

A consequence of analyzing \tit{mee} `with' as a postposition is that \tit{r} and \tit{mee} always form a constituent to the exclusion of \tit{w} and \tit{a}. At first sight this seems problematic, because it is possible for \tit{waar} `where', stranding \tit{mee} `with'. I repeat the relevant example in \ref{ex:meestranded}.

@@

The described pattern follows from a core assumptions in Nanosyntax: phrasal spellout spells out constituents. Looking more into detail, \tit{waarmee} `with what' takes precedence over \tit{met wat} `with what' because \tit{mee} has the structure of a postposition. Following the spellout algorithm, \tit{'t} `it' is replaced by \tit{'r} `there', so that it can combine with \tit{mee} `with'. The preposition \tit{met} `with' is only used if there is no other option to spell out the features.
